---
title: "difPlot - Visualizing DIF Analysis"
author: "Pavel Panko"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Vignette Title}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

The difPlot function is a flexible utility used to create publication-quality graphics for examining Differential Item Functioning in dichotomous item models. This function also provides several useful utilities to facilitate graphic creation. The general flow of tasks within the function is as follows: 

- Use the `ltm` package to estimate a separate model between the chosen groups.  
- Calculate the probability using the estimated parameters and supplied theta values for each group. 
- Plot the resultant probability against the theta values using `ggplot2`. 
- Press Enter to toggle between plots.
- Save the plot if desired. 

<a name = "top"/>

## Getting Started

#### Essential
difPlot is part of the irtPlot package available on github. To download the package, install the `devtools` package and run the following lines:

```{r, message = FALSE}
library(devtools)
install_github("ppanko/irtPlot")
```
Before using the function, the user must provide three arguments. First, the user must declare a data set for which he or she wishes to create graphics. Second, a sequence of theta values should be selected in order to plot the parameters. Lastly, the grouping variable should be declared as a factor. In the example below, the data are drawn from `difR` package and the theta values are chosen accordingly using the `seq` function, while the grouping variable is set to "Gender". 
```{r, message = FALSE}
library(irtPlot)
library(difR)

data(verbal)
data <- verbal[,1:2]
group <- factor(verbal[,"Gender"]) 
levels(group) <- c("Male", "Female")
theta <- seq(-3,3, 0.01)
```
Next, the user must choose which `model` and `type` to plot. This function currently only supports `"1PL"` and `"2PL"` models with `"3PL"` in development. Plot type has one option, [Item Characterstic Curves] (#icc)`"icc"`. See relevant sections below for more information on types. 


#### Options

Having dispensed with the required arguments, the user is free to select a number of optional commands to personalize the desired plot. Most of the the arguments in this section are used to tweak default setting for the `save` utility. 

- *ddir*: A character string specifying the directory in which to save plots. Default is the current working directory. 

- *save*: A logical value specifying whether or not to save the resultant graphics. Requires `TRUE/FALSE`; default is `FALSE`. 

- *title*: A character string to specify the name of the plot. Default behavior is to specify the plot type and the number of items in the model. 

- *filename*: A character string indicating the desired filename for the saved plot. The default filename is based on the model and type arguments. 

- *dpi*: An integer value specifying the resolution of the saved graphic. Default value is 300; recommended value for standard poster-size figures is 600. 

- *width*: A numeric value for the desired width of the figure in inches. Default is 8.5 inches.

- *height*: A numeric value for the desired height of the figure in inches. Default is 10 inches. 

- *itmNam*: A vector of character strings specifying the name of the items displayed in the legend. Default values are the names of the columns in the specified data set. 

- *silent*: A logical value specifying whether or not to print graphics to screen. This feature is primarily used when the user simply wants to save a plot without examining it. Default is `FALSE`. 

The user is encouraged to view the help files using the `?difPlot` function. 

<a name = "icc"/>

## I. Item & Test Charactersitic Curves 

For dichotomous items, Item Characteristic Curves are used to model the probability of answering a given item correctly across a range of ability values. The 3PL probability of a correct answer for subject *s* on item *i* is given by the following formula: 

$$P(Y_{is}=1|\theta_s) = c_i + (1 - c_i)\frac{exp(1.7a_i(\theta_s - b_i))}{1 + exp(1.7a_i(\theta_s - b_i))}$$

Examples the difference in the Item Characteristic Curves between Gender for `"2PL"` model of the `verbal` items are shown below. 

[Back to top](#top) \\
[Next Section](#ref)

***

```{r, fig.height=4.5, fig.width=5.5, fig.align = 'center', dev = 'svg', warning = FALSE}
difPlot(data, theta, grp = group, model = "2PL", itmNam = c("Curse", "Scold"))
```

***

<a name = "ref"/>

## References 
Embertson, S.E., & Reise, S. P. (2000). *Item Response Theory for Psychologists*. Hillsdale, NJ: Erlbaum.

Lee, J. (2016). *Lectures on Item Response Theory*. Personal Collection of J. Lee, Texas Tech University, 
  Lubbock, TX.

Magis, D., Beland, S., & Raiche, G. (2015). difR: Collection of methods to detect dichotomous differential item functioning (DIF). R package version 4.6

Rizopoulos D. (2006). ltm: An R package for Latent Variable Modelling and Item Response Theory Analyses. 
  *Journal of Statistical Software, 17*(5), 1-25. URL http://www.jstatsoft.org/v17/i05/

Wickham, H. (2009). *ggplot2: Elegant Graphics for Data Analysis*. New York: Springer-Verlag.

### See Also
- irtPlot
- polyPlot

***
[Back to top](#top)

Please leave comments at the project repository: https://www.github.com/ppanko/irtPlot 

