---
title: "irtPlot - Plotting Binary Items"
author: "Pavel Panko"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Vignette Title}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

Vignettes are long form documentation commonly included in packages. Because they are part of the distribution of the package, they need to be as compact as possible. The `html_vignette` output type provides a custom style sheet (and tweaks some options) to ensure that the resulting html is as small as possible. The `html_vignette` format:

- Never uses retina figures
- Has a smaller default figure size
- Uses a custom CSS stylesheet instead of the default Twitter Bootstrap style

## Getting Started

Note the various macros within the `vignette` section of the metadata block above. These are required in order to
```{r, message = FALSE}
library(irtPlot)
data <- LSAT
theta <- seq(-6,3, 0.01)
```

instruct R how to build the vignette. Note that you should change the `title` field and the `\VignetteIndexEntry` to match the title of your vignette.

## Item & Test Charactersitic Curves 

The `html_vignette` template includes a basic CSS theme. To override this theme you can specify your own CSS in the document metadata as follows:

```{r, fig.height=4.5, fig.width=5.5, fig.align = 'center', dev = 'svg'}
irtPlot(data, theta, model = "2PL", type = "icc")
irtPlot(data, theta, model = "2PL", type = "tcc")
```

## Item & Test Characteristic Curves 

The figure sizes have been customised so that you can easily put two images side-by-side. 

```{r, fig.height=4.5, fig.width=5.5, fig.align = 'center', dev = 'svg'}
irtPlot(data, theta, model = "3PL", type = "iif")
irtPlot(data, theta, model = "3PL", type = "tif")
```

## Likelihood & Log-Likelihood 

The figure sizes have been customised so that you can easily put two images side-by-side.

```{r, fig.height=4.5, fig.width=5.5, fig.align = 'center', dev = 'svg'}
irtPlot(data, theta, model = "1PL", type = "likl", subS = data[1,])
irtPlot(data, theta, model = "1PL", type = "logl", subS = data[1,])
```

You can enable figure captions by `fig_caption: yes` in YAML:

    output:
      rmarkdown::html_vignette:
        fig_caption: yes

Then you can use the chunk option `fig.cap = "Your figure caption."` in **knitr**.

## More Examples

You can write math expressions, e.g. $Y = X\beta + \epsilon$, footnotes^[A footnote here.], and tables, e.g. using `knitr::kable()`.

```{r, echo=FALSE, results='asis'}
knitr::kable(head(mtcars, 10))
```

Also a quote using `>`:

> "He who gives up [code] safety for [code] speed deserves neither."
([via](https://twitter.com/hadleywickham/status/504368538874703872))
